```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# **Set up HPC**


## **Access Network**

Ensure the device’s connection is on the VAI network, and the user enters the VAI username and password if needed, the VAI_guest network will not give users access to the VAI server.


## **MAC User**

Log into the terminal with **ssh vai_username@access.hpc.vai.org**, and enter the VAI password.


## **Window User**

If MobaXterm is not installed, please download the [free MobaXterm version](https://mobaxterm.mobatek.net/download.html), then set up, and agree with license agreements to process installation.


- Open MobaXterm
- Click on `Session` in the top left corner
- Click on `SSH`(Secure Shell)
- Enter `access.hpc.vai.org` in Remote host*
- Click the little box on the left of Specify username, then type VAI `username` and enter VAI `password`
- Next, click `OK`

Once the user successfully login, the user’s command prompt should look similar to `[username@submit001 ~]$` or `[username@submit002 ~]$`
<br><br>

# **File Navigation Exercise 1**

Practice directory: `/varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2`. This directory allows users to practice, and familiarize some basic Linux command lines.  Follow each command and review the solution below.

## **Navigate to the `EX2` Directory**
`Ex2` directory is inside of the `HPC_mini_workshop` directory, evidently, it requires to access `varidata` directory and a few directories before reaching the `Ex2` directory.

```{bash, eval = FALSE, engine= "sh"}
cd /varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2
```

## **Location the Current Directory**

```{bash, eval= TRUE, engine= "sh"}
pwd
```
## **View Inside of the `Ex2` Directory**

```{bash, eval= FALSE, engine= "sh"}
ls
```

## **Navigate to Home Directory**

```{bash, eval = FALSE, engine= "sh"}
cd
```

## **Create a Directory**
`hpc_mini_workshop` will be the directory name

```{bash, eval = FALSE, engine= "sh"}
mkdir hpc_mini_workshop
```

## **Copy a File**
Access multiple directories to reach `test_54_R1.fq` file and copy it to the `hpc_mini_workshop` directory.

```{bash, eval = FALSE, engine= "sh"}
cp /varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2/test_54_R1.fq hpc_mini_workshop
```

## **Check the Existing File**

```{bash, eval= FALSE, engine= "sh"}
cd hpc_mini_workshop

ls
```

## **Search a Pattern**

```{bash, eval= TRUE, engine= "sh"}
grep "AATTGG" test_54_R1.fq
```

## **Display Contents**

```{bash, eval= TRUE, engine= "sh"}
head test_54_R1.fq
```

## **Display the Number of words, Lines, and Characters**

```{bash, eval= TRUE, engine= "sh"}
wc test_54_R1.fq
```

## **Create a File**

```{bash, eval= TRUE, engine= "sh"}
touch exampleFile.txt
```

# **Pipe Command Line**

Pipe is a neat structure to process individual commands with a vertical pipe character between each command such as `command 1| command 2 | command 3`. According to Rahul Awati, “The pipe provides the connection that transforms an input into an output stream, while the filter reads from the standard input, does some processing on what it reads and writes to standard output.”

```{r, echo= FALSE}
library(knitr)
library(kableExtra)

# Create a data frame
df <- data.frame(
  Example = c("1", "2", "3"),
  Command = c("`cat file.txt | head -n7`", "`cat file.txt | head -n7 | tail -n5`", "`cat file.txt | head -n7 | tail -n5 | grep 'CGAG'`"),
  Description = c(
    "Display the contents of `file.txt` by selecting the first 7 lines. ",
    "Display the contents of `file.txt` by selecting the last 5 lines from the first 7 lines",
    "Display contents of `file.txt` by selecting the first 7 lines, then the last 5 lines, and finally filter those lines and display only lines with the pattern `CGAG`."
  )
)

# Render the data frame as a table in RMarkdown
kable(df, format = "html") %>%
  kable_styling() %>%
  column_spec(3, width = "40%") %>%
  row_spec(2, background = "#e9f5f8")

```

## **File Navigation Exercise 2**

In this exercise, we will continue to practice and advance the skill to access `/varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2` directory with character pipe `|`.

## **Navigate to the Home Directory**

```{bash, eval = FALSE, engine= "sh"}
cd
```

## **Display the Ex2 Directory Contents**

```{bash, eval = TRUE, engine= "sh"}
ls /varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2/
```

## **Copy Files**

Previous step, we can spot `protein_seq.fa` file and `test_54_R2.fq` file inside the `Ex2` directory. Copy the following code to the personal `hpc_mini_workshop` directory below.

```{bash, eval = FALSE, engine= "sh"}
cp /varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2/test_54_R2.fq hpc_mini_workshop

cp /varidata/researchtemp/hpctmp/HPC_mini_workshop/Ex2/protein_seq.fq hpc_mini_workshop
```

## **Verify Files are in the Directory**

```{bash, eval =TRUE, engine= "sh"}
cd hpc_mini_workshop

ls
```

## **Practice Concatenate**

Display the number of words, lines, and characters in the `protein_seq.fa` file.
```{bash, eval= TRUE, engine= "sh"}
cat protein_seq.fa | wc
```

Display the number of words in the `protein_seq.fa` file.
```{bash, eval= TRUE, engine= "sh"}
cat protein_seq.fa | wc -l
```

## **Find patterns**

Only display the beginning of 20 lines from `test_54_R2.fq` file with pattern **"AGCA"**
```{bash, eval= TRUE, engine= "sh"}
cat test_54_R2.fq | head -n20 | grep "AGCA"
```

## **Combine Files**

This step requires output redirection `>`. Output redirection allows users to perform specific commands and direct to preferred files. For example `cat protein.fq | head -n5 > first_5_protein.fq`, the command is displaying `protein.fq` file contents with 5 lines from the beginning of the file, then those 5 lines are redirected to a new file named `first_5_protein.fq`. Follow the command and combine two files into a single file below.

The `protein_seq.fa` file and the `test_54_R2.fq` file are combined into a single file called `combinedFile.fq`.
```{bash, eval=  TRUE, engine= "sh"}
cat protein_seq.fa | cat test_54_R2.fq > combinedFile.fq
```

## **Ensure Files Are In The Directory**

```{bash, eval= FALSE, engine= "sh"}
ls
```

## **Display the Contents From the End of the File**

```{bash, eval= TRUE, engine= "sh"}
cat combinedFile.fq | tail -n20
```
