
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Mini Project**

## **Starting an Interactive Job**

Now we have a better understanding of what Linux command lines are and how to inspect and understand the file system. In this mini project, we will make a transition from computationally intensive operations on the submit node to preallocated computational resources.

It is important to use preallocated computational resources to activate a job, the command prompt will switch from a submission prompt to a running prompt. For example,`[username@submit001 ~] $` to `[username@compute001 ~] $`.


Let’s begin by activating a job, it requires a number of nodes and time based on the user’s preference.
```{bash, eval= FALSE, engine= "sh"}
srun --nodes=1 --ntasks-per-node=1 --time=01:00:00 --pty bash -i
```


After a job has been submitted, users can verify how much time has been passed by coding.
```{bash, eval= FALSE, engine= "sh"}
squeue -u username
```

## **Create Directory**

Create a project directory with a **personal username** and a sub-directory for storing raw sequence data file name **fastqs** in `cd v/varidata/researchtemp/hpctmp/HPC_mini_workshop/Part3 directory`.

* Navigate to the `Part3` directory
```{bash, eval=FALSE, engine= "sh"}
cd /varidata/researchtemp/hpctmp/HPC_mini_workshop/Part3
```
* Create username directory
```
mkdir first.lastName
```
* Create a `fastqs` directory inside of the username directory

* Navigate to the username directory
```{bash, eval=FALSE, engine= "sh"}
cd first.lastName
```

* Create a `fastqs` directory
```{bash, eval= FALSE, engine= "sh"}
mkdir fastqs
```
* Ensure a directory is Exist
```{bash, eval=FALSE, engine= "sh"}
ls
```

## **Copy Files**

Verify the current directory is in the username directory, and copy an individual code below to receive the `md5sum.txt` file and a few `*fastq.gz` files.

```{bash, eval=FALSE, engine= "sh"}
cp /varidata/researchtemp/hpctmp/BBC/hpc_workshop_fqs/md5sum.txt ./fastqs/

cp /varidata/researchtemp/hpctmp/BBC/hpc_workshop_fqs/*fastq.gz ./fastqs/
```

## **Verify File**

Verify that `md5sum.txt` is transferred successfully by comparing it with `md5sums`. Copy the following code and it should return with **OK** for each file.

```{bash, eval=TRUE, engine="sh"}
cd fastqs

md5sum -c md5sum.txt
```

## **Back to the Username Directory**

Direct back to the username directory by entering the following.
```{bash, eval=FALSE, engine="sh"}
cd ..
```

## **Discover Compressed Zip File**

Notice all of the sequence data files ended with a `.gz` extension. A file with `.gz`(gzip) means that the file is compressed. If someone is interested to view the contents by using `cat`, the person will not be able to read it since it is not in a human-readable format. To view a `file.gz`, users require to use `zcat`.


Let’s jump into two sequence data files such as `SRR1039520_1.fastq.gz` and `SRR1039520_2.fastq.gz` which are in the `fastqs` directory.
```{bash, eval=TRUE, engine="sh"}
zcat fastqs/SRR1039520_1.fastq.gz | head
```

```{bash, eval=TRUE, engine="sh"}
zcat fastqs/SRR1039520_2.fastq.gz | head
```
Notice both `SRR1039520_1.fastq.gz` and `SRR1039520_2.fastq.gz` files have the exact read IDs.

## **Use `wc` Command Line to Count Lines and Characters**

Sequence data files tend to be large, however, it can be verified on how many lines of sequence it contains with `-l` and how many characters it contains with `-m`. Copy the code and view the result below.

```{bash, eval=TRUE, engine="sh"}
zcat fastqs/SRR1039520_1.fastq.gz | wc -l
```

The command below is retrieving data from `SRR1039520_1.fastq.gz` which is in `fastqs` directory, then receiving two lines from the beginning of the file and only taking the last line from the two lines that has been received. Finally, it counts the total number of characters in the last line.

```{bash, eval=TRUE, engine="sh"}
zcat fastqs/SRR1039520_1.fastq.gz | head -n2 | tail -n1 | wc -m

```

## **Working With Module**

There are many modules installed and available in `module av`, type `module av` to discover all the options. A smooth way to view and locate a portion of `modulefiles` is to construct pipe `|` after `module av 2>&1`.  `2>&1` is a redirection operator that redirects the error output of `module av`,  adding `head`, `grep` or other basic commands will help users view the `module` easier.

This command will output 50 lines of `modulefiles`
```{bash, eval=TRUE, engine="sh"}
module av 2>&1 | head -n50
```


This command will only output `bbc2` installed modules
```{bash, eval=TRUE, engine="sh"}
module av bbc2 2>&1 | head -n50
```


This command will output any modules containing the `fastqc` pattern
```{bash, eval=TRUE, engine="sh"}
module av bbc2 2>&1 | grep 'fastqc'
```

## **Create a Sequence Data File**

Create a `fastqc` directory inside of the username directory, this directory will process the rest of the sequence data files.  Each sequence data file will contain an HTML file once `fastqc` is output, users are free to view and have a better understanding of basic QC metrics which will be available on a regular browser.  Follow each command below and process

```{bash, eval=TRUE, engine="sh"}
module load bbc2/fastqc/fastqc-0.12.1

mkdir fastqc

# NOTE: Type 'fastqc -h' for more option to view the file
fastqc -o fastqc fastqs/*fastq.gz
```

Review Files
```{bash, eval=TRUE, engine="sh"}
ls fastqc

```

## **Set up a Job Script**

Now users have a better understanding of how to start an interactive job. If the interactive job time frame has not expired, please type `exit` to escape.  The next task will focus on submitting a non-interactive job since this is a practice, however, users can submit an interactive job.  Once users exit the interactive job, the command prompt will switch to the submission prompt as `[username@submit001 ~]$`.

Below provided a job script, please follow the commands below and set up a job script. Although the job script looks confusing, in the meantime, do not worry about how the code works.

```{bash, eval=FALSE, engine="sh"}
# Type ‘exit’ if an interactive job is active
exit

# Verify if an interactive job is active or expired
squeue -u username

# Go back to the project directory
cd /varidata/researchtemp/hpctmp/HPC_mini_workshop/Part3/firstname.lastname

```
## **Salmon Script**

This is what a Salmon Script looks like below, currently, do not worry about how the script works.
```{bash, eval=FALSE, engine="sh"}
#!/bin/bash

#SBATCH --nodes=1
#SBATCH --ntasks-per-node=1
#SBATCH --cpus-per-task=4
#SBATCH --time=00:02:00
#SBATCH --job-name=salmon
#SBATCH -e slurm-%j.err
#SBATCH -o slurm-%j.out

set -euo pipefail # see explanation at https://gist.github.com/mohanpedala/1e2ff5661761d3abd0385e8223e16425?permalink_comment_id=3945021

start_time=$(date +"%T")

# You need to navigate to your project directory. Conveniently, the $PBS_O_WORKDIR variable stores the path for where the job was submitted.
cd ${SLURM_SUBMIT_DIR}

module load bbc2/salmon/salmon-1.10.0

# Typically, you would have to first build an index before doing the aligning, but we have done this for you already. Here, we store the path to the index file in a variable called 'salmon_idx'.
salmon_idx="/varidata/research/projects/bbc/versioned_references/2022-03-08_14.47.50_v9/data/hg38_gencode/indexes/salmon/hg38_gencode/"

# make output directory for salmon
mkdir -p salmon

# this is called a for loop. We use this to run salmon quant on all the samples, one at a time. It is more efficient to run salmon on each sample "in parallel" but we will not do not today.
for samp_id in SRR1039520 SRR1039521
do
    salmon quant -p ${SLURM_NTASKS} -l A -i $salmon_idx -1 fastqs/${samp_id}_1.fastq.gz -2 fastqs/${samp_id}_2.fastq.gz -o salmon/${samp_id} --validateMappings

done

end_time=$(date +"%T")
echo "Start time: $start_time"
echo "End time: $end_time"
```


## **Submit a Job**

Once `run_salmon.sh` appears in the directory, the job script is ready to submit. To submit a job, require the `sbatch` command to submit the job. After the job is submitted, it will return a result similar to the one below with an ID.

```{bash, eval=FALSE, engine="sh"}
sbatch run_salmon.sh

```
Users can check if there is any task running in the background with the following command.  The job should take about two minutes to complete.
```{bash, eval=FALSE, engine="sh"}
squeue -u username

```

## **Process Standard Output(stdout) and Standard Error(stderr)**

After completing a job, the directory sorts out any standard error(stderr) and standard output(stdout) messages.  A file with **.err** is stderr and a file with **.out** is stdout.  Use `-ls` to check if there is any stderr and stdout in the directory.

```{bash, eval=FALSE, engine="sh"}
ls

```

View and process stderr and stdout results with the `tail` command.  As you notice, your stderr and stdout might look different from here.
```{bash, eval=TRUE, engine="sh"}
tail slurm-32564.err
```

Check process time.
```{bash, eval=TRUE, engine="sh"}
tail slurm-32564.out

```

## **View Transcript ID**

In this process, users will discover a specific gene called MUC1(protein) in the quant.sf file, and better understand how TPM(Transcripts per Million) values are stored in one of the files.

```{bash, eval=TRUE, engine="sh"}
head salmon/SRR1039520/quant.sf

```

The image shows that TPM is in the 4th column above, and transcript IDs are in Ensembl format. Next, use the grep command to locate ENST00000620103 canonical transcript in MUC1, see [Ensembl](http://useast.ensembl.org/Homo_sapiens/Gene/Summary?db=core;g=ENSG00000185499;r=1:155185824-155192916) for more detail.

Search `'ENST00000620103'` pattern in `salmon/SRR1039520/quant.sf` file
```{bash, eval=TRUE, engine="sh"}
grep 'ENST00000620103' salmon/SRR1039520/quant.sf

```

## **Use an interactive job to run multiQC on the Salmon and FastQC output**

In this final step, users will start an interactive job to perform multiQC to collect and summarize the outcomes from FastQC and Salmon, then evaluate the quality of the fastq sequences with the reference transcriptome(mapping rate).

Start an interactive job

```{bash, eval=FALSE, engine="sh"}
srun --nodes=1 --ntasks-per-node=1 --time=00:30:00 --pty bash -i

```

Negative to the project directory
```{bash, eval=FALSE, engine="sh"}
cd /varidata/researchtemp/hpctmp/HPC_mini_workshop/Part3/firstname.lastname/
```

Load BBC module
```{bash, eval=TRUE, engine="sh"}
module load bbc2/multiqc/multiqc-1.14
```

Multiqc creates a multiqc directory automatically, ensuring the period is after a space.
```{bash, eval=FALSE, engine="sh"}
multiqc --outdir multiqc .

```

View inside of the multiqc directory.
```{bash, eval=TRUE, engine="sh"}
ls multiqc
```

Now, the `multiqc_report.html` file is available to view or make a copy to a local directory.

# **Glossary**

```{r bash_cheatsheet, results='asis', echo=FALSE}
library(kableExtra)
library(knitr)

commands <- data.frame(
  Name = c(
    "Print Working Directory",
    "List",
    "List More Detail",
    "Make Directory",
    "Move",
    "Change Directory",
    "Remove",
    "Copy",
    "Search for File",
    "Head",
    "Tail",
    "Less",
    "More",
    "Quit",
    "Concatenate",
    "Search for Text",
    "Word, Line, and Character Count",
    "Touch"),

  Command_Line = c(
    "`pwd`",
    "`ls`",
    "`ls -lht`",
    "`mkdir`",
    "`mv`",
    "`cd`",
    "`rm`",
    "`cp`",
    "`find`",
    "`head`",
    "`tail`",
    "`less`",
    "`more`",
    "`q`",
    "`cat`",
    "`grep`",
    "`wc`",
    "`touch`"),

  Description = c(
    "Displays the current working directory",
    "Lists the files and directories in the current directory",
    "Display more details about the file",
    "Creates a new directory",
    "Moves or renames files or directories",
    "Change to an existing directory",
    "Deletes files and directories",
    "Copies files or directories",
    "Search for files and directories based on various criteria like name, size, and modification time",
    "Display at the beginning of a file",
    "Display at the end of a file",
    "Load the necessary portion of a file",
    "Load the entire file",
    "Stop viewing the current file",
    "Display the contents of a file",
    "Search for a specific pattern of text within files",
    "Display the number of words, lines, and characters in a file",
    "Create a new empty file"),

 Example = c(
  # explanation_1
  "**`[username\\@submit002 ~]$ pwd`**\n\nResult:/home/username",


  # explanation_2

  "**`[username\\@submit002 ~]$ ls`**\n\nResult: It returns empty after the $
  symbol since nothing has been created.",

 " **`[username\\@submit002 ~]$ ls -lht`**\n\nResult: Display file detail in the current director",

  # explanation_3
  "**`[username\\@submit002 ~]$ mkdir hpc_mini_workshop`**\n\nResult: A hpc_mini_workshop folder is created.",

  # explanation_4
  "**`[username\\@submit002 ~]$ mv hpc_mini_workshop workshopTraining`**\n\nResult: Now the hpc_mini_workshop directory is called workshopTraining",

  # explanation_5
  "**`[username\\@submit002 ~]$ cd workshopTraining`**\n\nResult: [username\\@submit002 workshopTraining]$ Notice ~ was in the home directory, now in the workshopTraining directory.",

  # explanation_6
  "**`[username\\@submit002 ~]$ rm -r TaskProject`**\n\n*Note: -r means directory\n\nResult: TaskProject is deleted",

  # explanation_7
  "**`[username\\@submit002 ~]$ cp -r Task1 Project`**\n\nResult: Task1 directory has moved to the Project directory.",

  # explanation_8
  "**`[username\\@submit002 ~]$ find Project`**\n\nResult: Task1 will appear within the Project directory\n\nProject\nProject/Task1",

  # explanation_9
  "**`[username\\@submit002 ~]$ head -n5 file_name`**\n\nResult: It will display the first 5 lines from the beginning of the file_name.",

  # explanation_10
  "**`[username\\@submit002 ~]$ tail -n5 file_name`**\n\nResult: It will display the last 5 lines from the end of the file_name",

  # explanation_11
  "**`[username\\@submit002 ~]$ less file.txt`**\n\nResult: The user is able to view a portion of the file.txt.",

  # explanation_12
  "**`[username\\@submit002 ~]$ more file.txt`**\n\nResult: The user is able to view the entire file.txt.",

  # explanation_13
  "quit viewing the current file",

  # explanation_14
  "**`[username\\@submit002 ~]$ cat file.txt`**\n\nResult: Display the contents of file.txt",

  # explanation_15
  "**`[username\\@submit002 ~]$ grep “GCGGA” sequence_file.fastq`**\n\nResult: Display any GCGGA pattern in sequence_file.fastq",

  # explanation_16
  "**`[username\\@submit002 ~]$ wc file.txt`**\n\nResult: Display the number of words, lines, and characters in the file.txt.",

  # explanation_17
  "**`[username\\@submit002 ~]$ touch exampleFile.txt`**\n\nResult: The command line will display file details in
  the current directory"
  ))

# Apply formatting to the Command_Line column
#commands$Command_Line <- cell_spec(commands$Command_Line, "html", font = "Courier", color = "black", background = "lightgray", escape = FALSE)

knitr::kable(
  commands,
  format = "html",
  col.names = c("Name", "Command Line", "Description", "Example"),
  align = c("l", "l", "l", "l"),
  escape = FALSE
) %>%
  kable_styling() %>%
  # Set background color for alternating rows
  row_spec(seq(2, nrow(commands), 2), background = "#e9f5f8")

```
